On finiteness of structures produced deterministically by co-transcriptional folding

### 2018-11-30 ttt.tex
update after noon
separate in subfiles
update 11/30

\documentclass[runningheads]{llncs}

\usepackage{tikz}
\usepackage{calc}
\usepackage{amsmath,amssymb}
\usetikzlibrary{automata}

\DeclareMathOperator*{\argmin}{arg\,min}

\usepackage{wasysym}

\begin{document}

\title{On finiteness of structures produced deterministically by co-transcriptional folding}

\section{Infiniteness of delay-1 unary deterministic oritatami system}

@@

\subsection{Introduction}

In this section, we consider the finiteness of structures produced deterministically at delay 1. Our result, cases of arity 1 and 3 can only yield finite structures of size $\mathcal{O} (n)$, and cases of arity 4 and more can only yield finite structures which is size of $mathcal{O} (n^2)$, and a case of arity 2 can yield infinite structures but they are only the zig-zag conformation shown in Fig.\ref{TTT_zigzag}.

@@

Let $\Xi$ be a deterministic oritatami system of delay 1 and arity 2. Assume its seed $\sigma$ consists of $n$ beads.
%Let us denote its transcript by $w = w_1 w_2 w_3 \cdots$ for some $w_1,w_2,w_3 \in \Sigma = \{e\}$.
For $i \geq 0$ let $C_i$ be the unique elongation of $\sigma$ by $w[1..i]$, that is, foldable by $\Xi$. Hence $C_0 = \sigma$.

Let us consider the stabilization of the $i$-th bead $a_i$ upon $C_{i-1}$. The bead cannot collaborate with any succeeding bead $w[i+1],w[i+2],\cdots$ at delay 1. There are just two ways to get stabilized at delay 1. One way is to be bound to another bead. The other way is through a $tunnel\ section$. A tunnel section consists of four beads that occupy four neighbors of a point (Fig.\ref{TTT_tunnel_intro}). 
Accordingly, how they are stabilized can be described by a binary sequence $S$ of $b$'s (bound) and $t$'s (tunnel section); priority is given to $t$, that is, $S[i] = t$ if the $i$-th bead $w_i$ is stabilized not only by being bound but also through a tunnel section.

@@

\begin{definition}[binding capability]
Let $B_i$ be $( \{ (h, i) | {}^\forall h < i \} \cup \{ (i, j) | {}^\forall j > i \}) \cap H$. Moreover, let $R_i$ be a set of neighbors of $w[i]$ that are free and reachable from $C_j$ where $C_j$ is a conformation which stabilized until $w[j]$.
 The number of binding capabilities of a conformation $C_j$ is denoted $\#bc(C_j)$ and is defined by $\sum^{j}_{k=-n+1} \min \{ |B_k|, |R_k|  \}$.
\end{definition}

@@

\begin{theorem}[Tunnel Troll Theorem]
Let $\Xi$ be a unary oritatami system of $\delta = 1, \alpha = 2$. If there are indices $i$ and $j$ such that $S[i..j+1] = bbt^{(j-i-1)}b$, then $\#bc(C_{i-1}) > \#bc(C_j)$ and if $S[i..j+1] = bt^lbt^mb \quad (l + m = j-i-1)$, then $\#bc(C_{i-1}) > \#bc(C_j)$. On the other hand, at $\delta = 1$ and $\alpha \geq 3$, if S$[k] = t$, then $\#bc(C_{k-1}) > \#bc(C_k)$.
\end{theorem}

@@

\proof%{Proof of Tunnel Troll Theorem}
Assume $\Xi$ is deterministic.
Each bead in the transcript is bound either inside a tunnel or outside. If a bead is stabilized inside a tunnel, then the position of successor is already decided either inside of a tunnel or outside. Moreover, if a bead is stabilized outside a tunnel, then its position is either an entrance of a tunnel or not.

Tunnel sections have three possible shapes up to symmetry : straight($A$), obtuse($B$) and acute($C$) turn (Fig. \ref{TTT_tunnel}), and we will consider each of those.

\begin{lemma}
\label{TTT_neighbor_lemma}
For unary transcripts at $\delta = 1$, if a bead has no free hand, then at least $\alpha + 2$ of its neighbors have to be occpied.
\end{lemma}

\begin{lemma}
\label{TTT_entrance_Tab}
Let $\Xi$ be an oritatami system at $\delta = 1, \alpha =2$. 
Assume $\Xi$ stabilizes the transcript until $w[i-1]$. If $w[i]$ is stabilized at an entrance point of tunnel $A$ or $B$, then $\#bc(C_{i-1}) > \#bc(C_{i})$.
\end{lemma}

\begin{lemma}
\label{TTT_exit}
Let $w[i]$ be a bead which is stabilized at the exit of a tunnel.
At $\delta = 1, \alpha =2$, if we assume $S[h..i+1] = bt^{(i-h)}b \quad (h<i)$, then $\#bc(C_{h-1}) \geq \#bc(C_i)$ and $\#bc(C_{i-2}) \geq \#bc(C_i)$.
%if we assume $\#bc(C_{i-2}) - \#bc(C_{i-1}) = a$, then $\#bc(C_{i}) - \#bc(C_{i-1}) \leq a$.
On the other hand, if we assume $S[k] = t (k \leq i)$ at $\delta =1, \alpha \geq 3$, then $\#bc(C_{k-1}) > \#bc(C_{k})$.
\end{lemma}

\begin{lemma}
\label{TTT_tunnelC_lemma}
Let $\Xi$ be a unary oritatami system of $\delta = 1, \alpha = 2$. We assume $S[h..i+1] = bt^{(i-h)}b \quad (1<h<i)$. If at least one of $w[h+1..i]$ is stabilized by tunnel $C$, then $\#bc(C_{h-3}) > \#bc(C_{h+1})$ and $\#bc(C_{h-3}) > \#bc(C_i)$.
\end{lemma}

Let us first consider cases of $\delta \geq 3, \alpha = 1$. These cases are clearly true because of lemma\ref{TTT_exit}.

Next, we consider the case of $\delta = 2, \alpha = 1$. We assume there is an index $h$ such that $S[h-1..h+1] = bbt$ or $S[h-1..h+1] = tbt$. According to lemma\ref{TTT_entrance_Tab}, if $w[h+1]$ is stabilized by tunnel $A$ or $B$, then $\#bc(C_{h-1}) > \#bc(C_{h})$. Also,  According to lemma\ref{TTT_tunnelC_lemma}, if $w[h+1]$ is stabilized by tunnel $C$, then $\#bc(C_{h-3}) > \#bc(C_{h+1})$.
On the other hand, if $S[k..l] = bt^{l-k}b$, then $\#bc(C_{k-1}) \geq \#bc(C_{l})$ because of lemma\ref{TTT_exit}. Therefore, if there are indices $i$ and $j$ such that $S[i..j+1] = bbt^{(j-i-1)}b$ or $S[i..j+1] = bt^mbt^nb \quad (m + n = j-i-1)$, then $\#bc(C_{i-1}) > \#bc(C_{j})$.

@@

          \node at (2,0) {Tunnel A};
        \end{tikzpicture}
      \end{minipage}

@@

          \node at (2,0) {Tunnel B};
        \end{tikzpicture}
      \end{minipage}

@@

          \node at (2,0) {Tunnel C};
        \end{tikzpicture}
      \end{minipage}

@@

\begin{proof}[lemma \ref{TTT_neighbor_lemma}]
Any transcript bead has predecessor and successor except for the first and last beads. If the bead does not have any free hand, then it uses hands with $\alpha$ neighbors. Thus, lemma \ref{TTT_neighbor_lemma} is clearly true.
\end{proof}

@@

\begin{itemize}
\item{Case of $t_0$}\\
  Let us consider points $n_3,n_4$. At least one of the points $n_3$ or $n_4$ is free because if both of them are occupied, $p^\prime$ is inside of tunnel. If $n_3$ is free, then $p^\prime$ has to be bound to a bead other than $n_1$ to deterministically stabilize. In this situation, at least three neighbors of $n_1$ are free, that is, $n_1$ has at least one free hand from lemma \ref{TTT_neighbor_lemma}. Hence, $p^\prime$ must be bound to $n_1$. Thus, a case of $t_0$ consumes two hands and it does not supply any binding capabilities.

@@

\begin{proof}[lemma \ref{TTT_exit}]
Fig.\ref{TTT_tunnel_exit} exhibits all the two kinds of exit of tunnel. 
At least one of points $n_1$ or $n_2$ is free because if both of them are occupied, $p^\prime$ is inside of tunnel.

\begin{paragraph}{}
$\delta = 1, \alpha = 2$\\
Let $\Xi$ be a unary oritatami system at $\delta = 1, \alpha = 2$. We assume $S[h..i+1] = bt^{(i-h)}b \quad (h<i)$.
Let $a$ be $\#bc(C_{i-2}) - \#bc(C_{i-1}) = a$. Then, $\#bc(C_{i}) - \#bc(C_{i-1}) \leq a$ as follows. Also, if $i - h > 1$ and $j$ is such that $h<j<i$, then $\#bc(C_{j-1}) \geq \#bc(C_{j})$ because all neighbors of $w[j]$ are occupied by beads forming the tunnel so that any $w[i+1..]$ cannot reach neighbors of $w[j]$. Thus, $\#bc(C_{h-1}) \geq \#bc(C_i)$ and $\#bc(C_{i-2}) \geq \#bc(C_i)$.
\end{paragraph}

@@

\begin{proof}[lemma \ref{TTT_tunnelC_lemma}]
Let $\Xi$ be a unary oritatami system at $\delta = 1, \alpha = 2$.
Assume $S[h..i+1] = bt^{i-h}b (h<i)$. If at least one of $w[h+1..i]$ are stabilized by tunnel $C$, then only $w[h+1]$ can use tunnel $C$ because if $w[g]$ which is one of $w[h+2..i]$, with $h+2 \leq g \leq i$ is stabilized by tunnel $C$, $C_g$ is a terminal.

Let us consider stabilization $S[h-1..h+1] = tbt$ or $S[h-1..h+1] = bbt$ as follows. In result, $\#bc(C_{h-3}) > \#bc(C_{h+1})$. In addition according to lemma\ref{TTT_exit} $\#bc(C_{h+1}) \geq \#bc(C_{i})$. Thus, $\#bc(C_{h-3}) > \#bc(C_{h+1})$ and $\#bc(C_{h-3}) > \#bc(C_{i})$.
\\

@@

\item{Right of Fig.\ref{TTT_tunnelC_enter_usingTunnel}}\\
  These cases are divided on number of capabilities that $w[h-1]$ consumes.
  \begin{itemize}
  \item[-]{$w[h-1]$ does not consume any bindings}\\
  According to lemma\ref{TTT_exit}, $\#bc(C_{h-3}) \geq \#bc(C_{h-1})$ because of $S[h-1] = t$.
    $w[h]$ has to bound one of $n_1, n_2, n_3$ in order to stabilize deterministically so that  $\#bc(C_{h-1}) > \#bc(C_{h})$.
    $w[h+1]$ has to be bound to $w[h-1]$ because $w[h-1]$ has bindings, that is, $w[h+1]$ consumes at least one hand and supplies at most one hand so that $\#bc(C_{h}) \geq \#bc(C_{h+1})$. Thus, in this cases $\#bc(C_{h-3}) > \#bc(C_{h+1})$.
%     This time, let us consider either $n_4$ is occupied or not. If $n_4$ is occupied, then $w[h-1]$ has no active bindings that is this situation consumes some binding capabilities. If $n_4$ is free and $w[h+2]$ is stabilized in $n_4$, then $w[h-1]$ has to bind $w[h+2]$. Therefore, In this case, stabilization of $w[h-1..h+2]$ consumes some bindings. If $n_4$ is free and $w[h+2]$ is stabilized except $n_4$, then this oritatami system has to use two binding capabilities in order to bind $w[h+2]$. Therefore, in this case consumes some bindings. Thus in this cases $\#bc(C_{h-1}) > \#bc(C_{h+2})$.

@@

\end{document}

>>>>

@@

In this section, we consider Problem~\ref{prob:det_unary_length} at delay 1. Our result, cases of arity 1 and 3 can only yield finite structures of size $O (n)$, and cases of arity 4 and more can only yield finite structures which is size of $O (n^2)$, and a case of arity 2 can yield infinite structures but they are only the zig-zag conformation shown in Fig.~\ref{TTT_zigzag}.

@@

%Let $\Xi$ be a deterministic oritatami system of delay 1 and arity 2. Assume its seed $\sigma$ consists of $n$ beads.
For $i \geq 0$ let $C_i$ be the unique elongation of $\sigma$ by $w[1..i]$, that is, foldable by $\Xi$. Hence $C_0 = \sigma$.

Let us consider the stabilization of the $i$-th bead $a_i$ upon $C_{i-1}$. The bead cannot collaborate with any succeeding bead $w[i+1],w[i+2],\cdots$ at delay 1. There are just two ways to get stabilized at delay 1. One way is to be bound to another bead. The other way is through a $tunnel\ section$. A tunnel section consists of four beads that occupy four neighbors of a point (Fig.\ref{TTT_tunnel_intro}). 
Accordingly, how they are stabilized can be described by a sequence $S$ of $b$'s (bound) and $t$'s (tunnel section); priority is given to $t$, that is, $S[i] = t$ if the $i$-th bead $w_i$ is stabilized not only by being bound but also through a tunnel section. 
We let $S$ take the value $\blacksquare$ for halt.

@@

\begin{definition}[binding capability]
Let $C = (P,w,H)$ be a conformation.
Let $B_k$ be $H \cap \{ (i,j) \mid \mbox{$i=k$ or $j=k$} \}$. Moreover, let $R_k$ be a set of neighbors of the point $P[k]$ that are free and reachable from $C$.
 The binding capability of $C$, denoted by $\#bc(C)$, is defined by $\sum^{|w|}_{k=1} \min \{ |B_k|, |R_k|  \}$.
\end{definition}

@@

%remark

\begin{theorem}[Tunnel Troll Theorem]
Let $\Xi$ be a deterministic unary oritatami system of delay $\delta = 1$. 
At arity $\alpha \geq 3$, if $S[k] = t$ and $S[k+1] \neq \blacksquare$, then $\#bc(C_{k-1}) > \#bc(C_k)$.
On the other hand, at arity $\alpha = 2$, if there are indices $i,j$ such that $S[i..j+1]$ is either $bbt^{(j-i-1)}b$ or $bt^\ell bt^{j-i-1-\ell}b$ for some $\ell$, then $\#bc(C_{i-1}) > \#bc(C_j)$.
\end{theorem}

@@

\proof%{Proof of Tunnel Troll Theorem}
Each bead in the transcript is bound either inside a tunnel or outside. If a bead is stabilized inside a tunnel, 
then it has at most one free neighbor, and hence its successor it to be stabilized there.
Moreover, if a bead is stabilized outside a tunnel, then its position is either an entrance of a tunnel or not.

Tunnel sections have three possible shapes up to symmetry : straight($S$), obtuse($O$) and acute($A$) turn (Fig. \ref{TTT_tunnel}), and we will consider each of those.

%\begin{lemma}
%\label{TTT_neighbor_lemma}
%For unary transcripts at $\delta = 1$, if a bead has no free hand, then at least $\alpha + 2$ of its neighbors have to be occpied.
%\end{lemma}

\begin{lemma}
\label{TTT_entrance_Tab}
Let $\Xi$ be a deterministic unary oritatami system at delay $\delta = 1$, arity $\alpha =2$. 
Assume $\Xi$ stabilizes the transcript until $w[i-1]$. If $w[i]$ is stabilized at an entrance point of tunnel $A$ or $B$, then $\#bc(C_{i-1}) > \#bc(C_{i})$.
\end{lemma}

\begin{lemma}
\label{TTT_exit}
Let $\Xi$ be a deterministic unary oritatami system of delay $\delta = 1$. 
Let $w[i]$ be a bead which is stabilized at the exit of a tunnel.
At $\delta = 1, \alpha =2$, if we assume $S[h..i+1] = bt^{(i-h)}b \quad (h<i)$, then $\#bc(C_{h-1}) \geq \#bc(C_i)$ and $\#bc(C_{i-2}) \geq \#bc(C_i)$.
%if we assume $\#bc(C_{i-2}) - \#bc(C_{i-1}) = a$, then $\#bc(C_{i}) - \#bc(C_{i-1}) \leq a$.
On the other hand, if we assume $S[k] = t (k \leq i)$ at $\delta =1, \alpha \geq 3$, then $\#bc(C_{k-1}) > \#bc(C_{k})$.
\end{lemma}

\begin{lemma}
\label{TTT_tunnelC_lemma}
Let $\Xi$ be a deterministic unary oritatami system of delay $\delta = 1$, arity $\alpha = 2$. We assume $S[h..i+1] = bt^{(i-h)}b \quad (1<h<i)$. If at least one of $w[h+1..i]$ is stabilized by tunnel $C$, then $\#bc(C_{h-3}) > \#bc(C_{h+1})$ and $\#bc(C_{h-3}) > \#bc(C_i)$.
\end{lemma}

Let us first consider cases of $\delta \geq 3, \alpha = 1$. These cases are clearly true because of Lemma~\ref{TTT_exit}.

Next, we consider the case of $\delta = 2, \alpha = 1$. We assume there is an index $h$ such that $S[h-1..h+1] = bbt$ or $S[h-1..h+1] = tbt$. According to Lemma~\ref{TTT_entrance_Tab}, if $w[h+1]$ is stabilized by tunnel $A$ or $B$, then $\#bc(C_{h-1}) > \#bc(C_{h})$. Also,  According to Lemma~\ref{TTT_tunnelC_lemma}, if $w[h+1]$ is stabilized by tunnel $C$, then $\#bc(C_{h-3}) > \#bc(C_{h+1})$.
On the other hand, if $S[k..l] = bt^{l-k}b$, then $\#bc(C_{k-1}) \geq \#bc(C_{l})$ because of Lemma~\ref{TTT_exit}. Therefore, if there are indices $i$ and $j$ such that $S[i..j+1] = bbt^{(j-i-1)}b$ or $S[i..j+1] = bt^mbt^nb \quad (m + n = j-i-1)$, then $\#bc(C_{i-1}) > \#bc(C_{j})$.

@@

          \node at (2,0) {Tunnel $S$};
        \end{tikzpicture}
      \end{minipage}

@@

          \node at (2,0) {Tunnel $O$};
        \end{tikzpicture}
      \end{minipage}

@@

          \node at (2,0) {Tunnel $A$};
        \end{tikzpicture}
      \end{minipage}

@@

@@

\begin{itemize}
\item{Case of $t_0$}\\
  Let us consider points $n_3,n_4$. At least one of the points $n_3$ or $n_4$ is free because if both of them are occupied, $p^\prime$ is inside of tunnel. If $n_3$ is free, then $p^\prime$ has to be bound to a bead other than $n_1$ to deterministically stabilize. In this situation, at least three neighbors of $n_1$ are free, that is, $n_1$ has at least one free hand from Lemma~ \ref{TTT_neighbor_lemma}. Hence, $p^\prime$ must be bound to $n_1$. Thus, a case of $t_0$ consumes two hands and it does not supply any binding capabilities.

@@

\begin{proof}[Lemma~ \ref{TTT_exit}]
Fig.\ref{TTT_tunnel_exit} exhibits all the two kinds of exit of tunnel. 
At least one of points $n_1$ or $n_2$ is free because if both of them are occupied, $p^\prime$ is inside of tunnel.

\begin{paragraph}{}
$\delta = 1, \alpha = 2$\\
Let $\Xi$ be a unary oritatami system at $\delta = 1, \alpha = 2$. We assume $S[h..i+1] = bt^{(i-h)}b \quad (h<i)$ and
let $a$ be $\#bc(C_{i-2}) - \#bc(C_{i-1}) = a$. Then, $\#bc(C_{i}) - \#bc(C_{i-1}) \leq a$ as follows. Also, if $i - h > 1$ and $j$ is such that $h<j<i$, then $\#bc(C_{j-1}) \geq \#bc(C_{j})$ because all neighbors of $w[j]$ are occupied by beads forming the tunnel so that any $w[i+1..]$ cannot reach neighbors of $w[j]$. Thus, $\#bc(C_{h-1}) \geq \#bc(C_i)$ and $\#bc(C_{i-2}) \geq \#bc(C_i)$.
\end{paragraph}

@@

\begin{proof}[Lemma~ \ref{TTT_tunnelC_lemma}]
Let $\Xi$ be a unary oritatami system at $\delta = 1, \alpha = 2$.
Assume $S[h..i+1] = bt^{i-h}b (h<i)$. If at least one of $w[h+1..i]$ are stabilized by tunnel $C$, then only $w[h+1]$ can use tunnel $C$ because if $w[g]$ which is one of $w[h+2..i]$, with $h+2 \leq g \leq i$ is stabilized by tunnel $C$, $C_g$ is a terminal.

Let us consider stabilization $S[h-1..h+1] = tbt$ or $S[h-1..h+1] = bbt$ as follows. In result, $\#bc(C_{h-3}) > \#bc(C_{h+1})$. In addition according to Lemma~\ref{TTT_exit} $\#bc(C_{h+1}) \geq \#bc(C_{i})$. Thus, $\#bc(C_{h-3}) > \#bc(C_{h+1})$ and $\#bc(C_{h-3}) > \#bc(C_{i})$.
\\

@@

\item{Right of Fig.\ref{TTT_tunnelC_enter_usingTunnel}}\\
  These cases are divided on number of capabilities that $w[h-1]$ consumes.
  \begin{itemize}
  \item[-]{$w[h-1]$ does not consume any bindings}\\
  According to Lemma~\ref{TTT_exit}, $\#bc(C_{h-3}) \geq \#bc(C_{h-1})$ because of $S[h-1] = t$.
    $w[h]$ has to bound one of $n_1, n_2, n_3$ in order to stabilize deterministically so that  $\#bc(C_{h-1}) > \#bc(C_{h})$.
    $w[h+1]$ has to be bound to $w[h-1]$ because $w[h-1]$ has bindings, that is, $w[h+1]$ consumes at least one hand and supplies at most one hand so that $\#bc(C_{h}) \geq \#bc(C_{h+1})$. Thus, in this cases $\#bc(C_{h-3}) > \#bc(C_{h+1})$.
%     This time, let us consider either $n_4$ is occupied or not. If $n_4$ is occupied, then $w[h-1]$ has no active bindings that is this situation consumes some binding capabilities. If $n_4$ is free and $w[h+2]$ is stabilized in $n_4$, then $w[h-1]$ has to bind $w[h+2]$. Therefore, In this case, stabilization of $w[h-1..h+2]$ consumes some bindings. If $n_4$ is free and $w[h+2]$ is stabilized except $n_4$, then this oritatami system has to use two binding capabilities in order to bind $w[h+2]$. Therefore, in this case consumes some bindings. Thus in this cases $\#bc(C_{h-1}) > \#bc(C_{h+2})$.

@@
